---
title: "Reproducible Research: Peer Assessment 1"
author: "Corbin Thornbrough"
date: "Sunday, August 17th, 2014"
output: html_document
---

##Loading and preprocessing the data

####The following code unzips the activity csv and reads it into R as the "activity" table.
```{r activity.read}
setwd("C:/Users/corbin.thornbrough/Documents/GitHub/RepData_PeerAssessment1")
unzip("activity.zip")
activity<-read.csv("activity.csv")
options(scipen=4)
```

####This code transforms the data types to make the data more usable.
```{r activity.transform}
activity$date<-as.Date(activity$date)
activity$steps<-as.numeric(activity$steps)
activity$interval<-as.numeric(activity$interval)
options(scipen=4)
```

##What is mean total number of steps taken per day?
```{r hist, echo=FALSE}
act.steps.agg<-aggregate(steps~date,activity,sum)
hist(act.steps.agg$steps, main="Histogram of Daily Steps", xlab="Steps")
```

```{r meanmedian1, results="hide", echo=FALSE}
mean1<-mean(act.steps.agg$steps)
median1<-median(act.steps.agg$steps)
```
The mean number of steps by day is `r mean1` while the median is `r median1`.

##What is the average daily activity pattern?
```{r interval.plot, echo=FALSE}
act.interval.agg<-aggregate(steps~interval,activity, mean)
plot(act.interval.agg$steps~act.interval.agg$interval, type="l", xlab="five-minute interval",
     ylab="steps averaged across all days", main="Plot of Average Steps by Interval")
```

```{r max.interval, echo=FALSE, results="hide"}
max<-act.interval.agg[which.max(act.interval.agg$steps),1]
```
The `r max` interval is the interval with the largest average number of steps.

##Imputing missing values

```{r count.na, echo=FALSE, results="hide"}
missing<-sum(is.na(activity$steps))
```
There are `r missing` missing values in the dataset.

####This creates a new dataset with NA values (for the steps variable) replaced by the mean number of steps for that interval over the week.
```{r activity.mod}
activity.mod<-activity
activity.mod$steps<-with(activity.mod,ifelse(
  is.na(steps),
  ave(steps,interval,FUN=function(steps)mean(steps, na.rm=TRUE)),
  steps
  ))
```

####Here is a histogram of the total number of steps taken each day with the NAs replaced.
```{r mod.hist, echo=FALSE}
mod.act.agg<-aggregate(steps~date,activity.mod,sum)
hist(mod.act.agg$steps, main="Histogram of Daily Steps", xlab="Steps")
```

```{r meanmedian2, echo=FALSE, results="hide"}
mean2<-mean(mod.act.agg$steps)
median2<-median(mod.act.agg$steps)
```

With the NAs replaced, the mean steps taken per day is `r mean2`, and the median steps taken per day is `r median2`. These are `r (mean2-mean1)/mean1`% and `r (median2-median1)/median1`% different, respectively. Obviously this implies that the meaning of the data isn't dramatically changed by adding mean values by interval back in to replace NA values.

##Are there differences in activity patterns between weekdays and weekends?

####This adds another variable to the dataset with NAs replaced.  This new variable denotes that the record was on a weekend (Saturday & Sunday) or a weekend (Monday-Friday).
```{r weekday}
activity.mod$weekday<-ifelse(weekdays(activity.mod$date)=="Saturday" | weekdays(activity.mod$date)=="Sunday","weekend","weekday")
```

####This plots the weekend and weekday number of steps (x-axis) by the 5-minute interval (y-axis).
```{r weekday.plot, echo=FALSE}
require(lattice)
xyplot(activity.mod$steps~activity.mod$interval | activity.mod$weekday, activity.mod, type="h", ylab="Steps",xlab="Interval")
```

As you can see, on the weekends, activity begins much later in the day and extends further into the evening than on the weekdays.